A tool for designing digital filters for the Hankel and Fourier transforms in potential, diffusive, and wavefield modeling

### 2018-06-19 ManuscriptGeophysics2018.tex
Notebooks ready for re-submission

% \documentclass[manuscript]{geophysics}
% \documentclass[paper,twocolumn,twoside]{geophysics}
% \documentclass[paper]{geophysics}
% \documentclass[paper, revised]{geophysics}
\documentclass[manuscript,revised]{geophysics}

@@

\subsection{Design}
Figure~\ref{fig:201} shows the solution spaces of \old{four}\new{eight}
consecutive inversion runs\old{, where each run is a focus on a subsection of
  the previous run, indicated by the red square. The filter length in this case
is $N=201$, and}\new{where} the theoretical transform pairs
\old{are} as given, e.g., in \cite{USGS.75.Anderson}\old{:}\new{,}
%
\begin{align}
  \int^\infty_0\,l \exp\left(-al^2\right) J_0(rl)\,dl &=
  \frac{\exp\left(\frac{-r^2}{4a}\right)}{2a}\ , \\
  \int^\infty_0\,l^2 \exp\left(-al^2\right) J_1(rl)\,dl &=
  \frac{r}{4a^2} \exp\left(-\frac{r^2}{4a}\right)\ ,
  \label{eq:j01}
\end{align}
%
%
\plot*{201}{width=\textwidth}{%
  Solution spaces of four consecutive inversion runs for a 201\,pt filter. Each
  consecutive run zooms into a portion of the previous solution space,
  indicated by the red square. The more in detail we obtain the solution space,
  the more random appears to be the distribution.}%
%
where \new{used; } $a$ was set to 5. In the algorithm you can provide one pair
for the inversion, and a different pair to get the minimum amplitude or the
maximum $r$. In this case the same transform pair was used for both the
inversion and the check of quality. The acceptable relative error was set to
1\,\%. The RHS evaluation parameter for the inversion was set so  that an
over-determined system was evaluated with $M = 2N$ equations, where $r_\mr{I}$
was logarithmically spaced from $\log_{10}(1/\max(b)) - 1$ to
$\log_{10}(1/\min(b)) + 1$, where $b$ are the filter abscissae as given in
equation~\ref{eq:base}. This is accomplished in \texttt{fdesign} by setting
parameter $r_\mr{def} = (1, 1, 2)$.

\new{The top row shows in subplots (a) to (d) four different inversion runs
for $N=51$, 101, 201, and 401 for a wide range of spacing and shift values.
From this one has to decide for a filter length. Longer filters tend to give
more precise results, but their transform is computationally more expensive as
more values have to be evaluated. The length of the filter does not has to be
an odd number, but many published filters do have an odd number of
coefficients. The integral runs from $-\infty$ to $+\infty$ and hence an odd
point filter comes naturally by truncating the integral to a finite sum from
$-(N-1)/2$ to $(N-1)/2$. However, this is only true if the shift value is
$\delta = 0$. If your shift value is not zero then there is no advantage in
having an odd point filter.}

\new{The bottom row shows four consecutive inversion runs for $N=201$, where
each run is a focus on a subsection of the previous run, indicated by the red
square.} From the low resolution overview runs (e) and (f) it looks like this
is a standard, straight-forward minimization problem. However, from the more
detailed results in (g) and (h) it becomes obvious that it is a minimization
problem that has to be solved stochastically, as there are solutions with 2
orders of magnitude difference apparently randomly next to each other.

Figure~\ref{fig:201b} shows in (a) the filter values for $J_0$ and $J_1$ of the
best 201 point filter obtained this way, and in (b) its right-hand-side
solution. The black dots indicate negative values, from which it can seen that
adjacent values are often alternating between positive and negative
contributions.
%
\plot*{201b}{width=.9\textwidth}{(a) Filter values of the best obtained 201\,pt
  filter with the corresponding check of quality in (b). Black points indicate
  negative values, which shows that adjacent values have often opposite signs.}
%

Designing filters is to a large extent trial and error. All input variables
influence the outcome, and often a good filter is found by sheer luck of
choosing the right starting parameters\new{, as can be seen in
Figure~\ref{fig:201b} (h). However, once you are in this zone every filter is a
very good filter}. Each of the input parameters has an effect on the outcome.
It also depends heavily on the transform pairs $f_\mr{I;C}$; functions that
decay rapidly are generally better, as noted by earlier authors
\citep[e.g.][]{USGS.75.Anderson}. It also depends on the filter length, and
obviously heavily on the spacing and shift values. Another important point is
how you define the right-hand evaluation points of the inversion
($r_\mr{def}$). Evaluating corresponding transform pairs separately or jointly
also leads to different filter coefficients ($J_0$, $J_1$, or $J_0$ \& $J_1$,
or equally sine, cosine, or sine \& cosine); also if the real or the imaginary
part is used when complex transform pairs or \texttt{empymod} is used; and if
it is inverted for the minimum amplitude or for the maximum $r$.

@@

Figure~\ref{fig:kongf5} compares the derived 201\,pt filter with the two
half-spaces case used by \cite{GP.07.Kong} in his Figure~5. The model consists
of a $0.3125\,\Omega\,$m sea layer of infinite thickness above a subsurface
$1\,\Omega\,$m half-space. The signal of a $1\,$Hz frequency $x$-directed
electric source 50\,m above the interface is measured at an $x$-directed
electric receiver at the interface.
%
\plot*{kongf5}{width=.9\textwidth}{(a) Results of different filters and QWE for
  the model of Figure~5 from \cite{GP.07.Kong} with the relative errors shown
  in (b), using the QWE result. The relative error is meaningless from about
  15.5\,km onwards, as QWE failed as well for these very low amplitudes.}
%
In (a) it can be seen that the new 201\,pt filter is able to recover smaller
amplitudes than the 241\,pt filter from \cite{GP.07.Kong}, the 201\,pt filter
from \cite{GEO.12.Key}, and the 801\,pt filter from \cite{TMS.82.Anderson}
(Wer201, Kong241, Key201, and And801, respectively). It behaves equally well as
the quadrature with extrapolation (QWE), for which we used a 51\,pt quadrature
with relative and absolute tolerance of $10^{-12}$ and $10^{-30}$,
respectively. The relative error is shown in (b), where the QWE result was
taken as \emph{truth}. For offsets greater than roughly 15.5\,km the relative
error becomes meaningless, as the QWE fails itself; this part is greyed out in
the figure.

Figure~\ref{fig:keyf5}(a) considers the canonical CSEM model as used in
\cite{GEO.12.Key} in his Figure~5: a $0.303\,\Omega\,$m sea with depth of
2\,km, a background resistivity of $1\,\Omega\,$m, in which a $100\,\Omega\,$m
target layer of 100\,m thickness is embedded at 1000\,m below the seafloor.
The source is located at 1990\,m and receivers are positioned on the seafloor.
%
\plot*{keyf5}{width=.9\textwidth}{(a) The canonical marine CSEM model of
  \cite{GEO.12.Key} and (b) a land model with source and receiver at the
  surface. The new filter has generally relative errors which are several
  orders of magnitude lower than the other filters.}
%

@@

It is very important to note again that other scenarios might yield very
different error plots. Although the new 201\,pt filter proves to be very
accurate for these three models, it might not be the best filter in other
cases. To verify the applicability of the filter we run two different tests
over a wide range of scenarios. Figure~\ref{fig:error} shows the relative
errors of DLF using the filter \emph{Wer201} compared to QWE for a deep water
model, a shallow water model, and a land model. The models consist of air
(upper halfspace), a water layer in the first two examples, and a subsurface
halfspace. Source depth is 1990\,m, 10\,m, and 0.5\,m in the deep, shallow, and
land case, respectively. Receiver depth is 2000\,m, 100\,m, and 0.8\,m. The
subsurface halfspace resistivity varies from 1 to 10, 100, and
1000\,$\Omega\,$m from the left to the right column. The error plots cover
offsets from 50\,m to 20\,km on the x-axis, and 0.01\,Hz to 10\,Hz on the
y-axis. It can be seen in the result that the filter is sufficiently precise
for all practical CSEM cases, hence has generally an error far below 1\,\%
($10^{-2}$) in all cases. The exception are large offsets with relatively high
frequencies, e.g., 10\,km offset with 8\,Hz. In the shallow water case this
zone has a very high error for subsurface resistivities of 1\,$\Omega\,$m and
10\,$\Omega\,$m (yellow zones). However, this is due to the fact that the
amplitudes of the signal for these offsets and frequencies are in the order of
$10^{-25}\,$V/m.
%
\plot*{error}{width=\textwidth}{Relative error plots comparing the DLF
  (\emph{Wer201}) with QWE for the three cases deep water (top row), shallow
  water (middle row), and land (bottom row). The subsurface halfspace
  resistivity varies from 1 to 1000\,$\Omega\,$m from the left column to the
  right column; offsets vary from 50\,m to 20\,km, and frequencies vary from
  0.01\,Hz to 10\,Hz. The DLF provides sufficiently accurate results for all
  practical CSEM applications. The exception is in the shallow water case for
  $\rho = 1\,\Omega\,$m and $10\,\Omega\,$m for large offsets and high
  frequencies (yellow zones). However, the amplitudes in these regions are in
  the order of $10^{-25}\,$V/m, and therefore \old{far}\new{several orders of
  magnitude} below the noise level\new{s of current EM instrument systems}.}
%

>>>>

% \documentclass[manuscript]{geophysics}
% \documentclass[paper,twocolumn,twoside]{geophysics}
\documentclass[paper]{geophysics}
% \documentclass[paper, revised]{geophysics}
% \documentclass[manuscript,revised]{geophysics}

@@

\subsection{Design}
\old{Figure~\ref{fig:201b} shows the solution spaces of four consecutive
inversion runs, where each run is a focus on a subsection of the previous run,
indicated by the red square. The filter length in this case is $N=201$, and the
theoretical transform pairs are as given, e.g., in
\cite{USGS.75.Anderson}:}
\new{Figures~\ref{fig:201a} and~\ref{fig:201b} show the solution spaces of
seven consecutive inversion runs using the following theoretical transform
pairs \citep[e.g.,][]{USGS.75.Anderson},}
%
\begin{align}
  \int^\infty_0\,l \exp\left(-al^2\right) J_0(rl)\,dl &=
  \frac{\exp\left(\frac{-r^2}{4a}\right)}{2a}\ , \\
  \int^\infty_0\,l^2 \exp\left(-al^2\right) J_1(rl)\,dl &=
  \frac{r}{4a^2} \exp\left(-\frac{r^2}{4a}\right)\ ,
  \label{eq:j01}
\end{align}
%
where $a$ was set to 5. In the algorithm you can provide one pair
for the inversion, and a different pair to get the minimum amplitude or the
maximum $r$. In this case the same transform pair was used for both the
inversion and the check of quality. The acceptable relative error was set to
1\,\%. The RHS evaluation parameter for the inversion was set so  that an
over-determined system was evaluated with $M = 2N$ equations, where $r_\mr{I}$
was logarithmically spaced from $\log_{10}[1/\max(b)] - 1$ to
$\log_{10}[1/\min(b)] + 1$, where $b$ are the filter abscissae as given in
equation~\ref{eq:base}. This is accomplished in \texttt{fdesign} by setting
parameter $r_\mr{def} = (1, 1, 2)$.

\new{Figure~\ref{fig:201a} shows three different inversion runs for $N=101$,
201, and 401 for a wide range of spacing and shift values.
From this one has to decide for a filter length. Longer filters tend to give
more precise results, but their transform is computationally more expensive as
more values have to be evaluated. The length of the filter does not has to be
an odd number, but many published filters do have an odd number of
coefficients. The integral runs from $-\infty$ to $+\infty$ and hence an odd
point filter comes naturally by truncating the integral to a finite sum from
$-(N-1)/2$ to $(N-1)/2$. However, this is only true if the shift value
$\delta = 0$. If it differs from zero then there is no advantage in having an
odd point filter.}
%
\plot*{201a}{width=\textwidth}{%
  \new{Test of different filter lengths. Filter length is a trade-off between
    precision and speed, as shorter filters are faster but less precise, and
  longer filters are more precise but computationally more expensive. The red
  square in subplot (b) is the starting point of Figure~\ref{fig:201b}.} }%
%

\new{Figure~\ref{fig:201b} shows four consecutive inversion runs for $N=201$,
where each run is a focus on a subsection of the previous run, indicated by the
red square.} From the low resolution overview runs (a) and (b) it looks like
this is a standard, straight-forward minimization problem. However, from the
more detailed results in (c) and (d) it becomes obvious that it is a
minimization problem that has to be solved stochastically, as there are
solutions with 2 orders of magnitude difference apparently randomly next to
each other.
%
\plot*{201b}{width=\textwidth}{%
  Solution spaces of four consecutive inversion runs for a 201\,pt filter. Each
  consecutive run zooms into a portion of the previous solution space,
  indicated by the red square. The more in detail we obtain the solution space,
  the more random appears to be the distribution.}%
%

Figure~\ref{fig:201c} shows in (a) the filter values for $J_0$ and $J_1$ of the
best 201 point filter obtained this way, and in (b) its right-hand-side
solution. The black dots indicate negative values, from which it can seen that
adjacent values are often alternating between positive and negative
contributions.
%
\plot*{201c}{width=.9\textwidth}{(a) Filter values of the best obtained 201\,pt
  filter with the corresponding check of quality in (b). Black points indicate
  negative values, which shows that adjacent values have often opposite signs.}
%

Designing filters is to a large extent trial and error. All input variables
influence the outcome, and often a \old{good}\new{even better} filter is found
by sheer luck of choosing the right starting parameters\new{, as can be seen in
Figure~\ref{fig:201c} (h). However, once you are in this zone every filter is a
very good filter}. Each of the input parameters has an effect on the outcome.
It also depends heavily on the transform pairs $f_\mr{I;C}$; functions that
decay rapidly are generally better, as noted by earlier authors
\citep[e.g.][]{USGS.75.Anderson}. It also depends on the filter length, and
obviously heavily on the spacing and shift values. Another important point is
how you define the right-hand evaluation points of the inversion
($r_\mr{def}$). Evaluating corresponding transform pairs separately or jointly
also leads to different filter coefficients ($J_0$, $J_1$, or $J_0$ \& $J_1$,
or equally sine, cosine, or sine \& cosine); also if the real or the imaginary
part is used when complex transform pairs or \texttt{empymod} is used; and if
it is inverted for the minimum amplitude or for the maximum $r$.

@@

Figure~\ref{fig:kongf5} compares the derived 201\,pt filter with the two
half-spaces case used by \cite{GP.07.Kong} in his Figure~5. The model consists
of a $0.3125\,\Omega\,$m sea layer of infinite thickness above a subsurface
$1\,\Omega\,$m half-space. The signal of a $1\,$Hz frequency $x$-directed
electric source 50\,m above the interface is measured at an $x$-directed
electric receiver at the interface.
%
\plot*{kongf5}{width=\textwidth}{(a) Results of different filters and QWE for
  the model of Figure~5 from \cite{GP.07.Kong} with the relative errors shown
  in (b), using the QWE result. The relative error is meaningless from about
  15.5\,km onwards, as QWE failed as well for these very low amplitudes.}
%
In (a) it can be seen that the new 201\,pt filter is able to recover smaller
amplitudes than the 241\,pt filter from \cite{GP.07.Kong}, the 201\,pt filter
from \cite{GEO.12.Key}, and the 801\,pt filter from \cite{TMS.82.Anderson}
(Wer201, Kong241, Key201, and And801, respectively). It behaves equally well as
the quadrature with extrapolation (QWE), for which we used a 51\,pt quadrature
with relative and absolute tolerance of $10^{-12}$ and $10^{-30}$,
respectively. The relative error is shown in (b), where the QWE result was
taken as \emph{truth}. For offsets greater than roughly 15.5\,km the relative
error becomes meaningless, as the QWE fails itself; this part is greyed out in
the figure.

Figure~\ref{fig:keyf5}(a) considers the canonical CSEM model as used in
\cite{GEO.12.Key} in his Figure~5: a $0.303\,\Omega\,$m sea with depth of
2\,km, a background resistivity of $1\,\Omega\,$m, in which a $100\,\Omega\,$m
target layer of 100\,m thickness is embedded at 1000\,m below the seafloor.
The source is located at 1990\,m and receivers are positioned on the seafloor.
%
\plot*{keyf5}{width=\textwidth}{(a) The canonical marine CSEM model of
  \cite{GEO.12.Key} and (b) a land model with source and receiver at the
  surface. The new filter has generally relative errors which are several
  orders of magnitude lower than the other filters.}
%

@@

It is very important to note again that other scenarios might yield very
different error plots. Although the new 201\,pt filter proves to be very
accurate for these three models, it might not be the best filter in other
cases. To verify the applicability of the filter we run two different tests
over a wide range of scenarios. Figure~\ref{fig:error} shows the relative
errors of DLF using the filter \emph{Wer201} compared to QWE for a deep water
model, a shallow water model, and a land model. The models consist of air
(upper halfspace), a water layer in the first two examples, and a subsurface
halfspace. Source depth is 1990\,m, 10\,m, and 0.5\,m in the deep, shallow, and
land case, respectively. Receiver depth is 2000\,m, 100\,m, and 0.8\,m. The
subsurface halfspace resistivity varies from 1 to 10, 100, and
1000\,$\Omega\,$m from the left to the right column. The error plots cover
offsets from 50\,m to 20\,km on the x-axis, and 0.01\,Hz to 10\,Hz on the
y-axis. It can be seen in the result that the filter is sufficiently precise
for all practical CSEM cases, hence has generally an error far below 1\,\%
($10^{-2}$) in all cases. \new{A relative noise of less than 1\,\% is generally
considered really good for real data.} The exception\new{s} are \new{in the
shallow water case at} large offsets with relatively high frequencies, e.g.,
10\,km offset with 8\,Hz\old{. In the shallow water case}\new{;} this zone has
a very high error for subsurface resistivities of 1\,$\Omega\,$m and
10\,$\Omega\,$m (yellow zones). However, this is due to the fact that the
amplitudes of the signal for these offsets and frequencies are in the order of
$10^{-25}\,$V/m. \new{The grey contour lines in the plot indicate the power of
  the amplitude of the signal.}
%
\plot*{error}{width=\textwidth}{Relative error plots comparing the DLF
  (\emph{Wer201}) with QWE for the three cases deep water (top row), shallow
  water (middle row), and land (bottom row). The subsurface halfspace
  resistivity varies from 1 to 1000\,$\Omega\,$m from the left column to the
  right column; offsets vary from 50\,m to 20\,km, and frequencies vary from
  0.01\,Hz to 10\,Hz. The DLF provides sufficiently accurate results for all
  practical CSEM applications. The exception is in the shallow water case for
  $\rho = 1\,\Omega\,$m and $10\,\Omega\,$m for large offsets and high
  frequencies (yellow zones). However, the amplitudes in these regions are in
  the order of $10^{-25}\,$V/m, and therefore \old{far}\new{several orders of
  magnitude} below the noise level\new{s of current EM instrument systems}.
  \new{Contour lines show the power of the amplitude of the signal.}}
%
